FAIR fission track analysis with geochron@home

### 2025-07-02 geochron@home/manuscript/paper.tex
Minor edit to manuscript
geochron@home

geochron@home serves four different purposes. It can be used (1) to
count fission tracks in `private mode', i.e. hidden from other users
on the internet; (2) to archive fission track images and counts for
inspection by other users; (3) to create tutorials for new students of
the fission track method; and (4) to generate crowd-sourced fission
track data, by serving randomly selected selections of images to
citizen scientists. We illustrate these four applications with
examples that demonstrate (1) geochron@home's ability to compare and
combine fission track counts for multiple users within a lab group;
(2) the use of the geochron@home archive in the peer review system;
(3) the effectiveness of even simple tutorials in teaching novice
users how to count fission tracks; and (4) the `wisdom of crowds' in
fission track identification.\medskip

@@

\item Upload the data to the geochron@home platform. geochron@home is
  a Django web-app featuring a user interface powered by the Leaflet
  library (JavaScript).  The database can be accessed via a Python API
  and a (more limited) web-based GUI. Accessing the API requires
  administrator privileges. Administrators can create projects,
  samples and grains; download data; and set the access rights of
  `ordinary' users.  Projects can be private or public. The source
  code and installation instructions for geochron@home are freely
  available over GitHub (see the Data Availability statement at the
  end of this paper). This allows fission track users to set up their
  own server. Alternatively, fission track labs can upload their data
  to the UCL server by contacting the corresponding author.

\item Analyse the images with a browser-based `virtual fission track
  microscope'. This virtual instrument has a simple user interface
  with controls to zoom, pan and focus in or out of the digital image
  stack. Depending on the permissions granted to the user by the
  administrator, the virtual microsocope offers a number of different
  options.  Entry level `ordinary' users are only allowed to count
  tracks by clicking within a pre-defined `region of interest'. In
  contrast, `superusers' are allowed to define their own regions of
  interest. Once the user is satisfied that they have counted all the
  fission tracks in a particular grain, they can submit the results to
  the server. They are then presented with a new set of images until
  all grains are counted.

@@

The annotations in the current tutorial pages were made by an
experienced fission track analyst (Andrew Carter). This basic tutorial
provides a quick and easy mechanism to train novice users in the art
of fission track analysis. The tutorial pages are in their infancy,
and offer only a limited degree of interactivity. Users can click on
features to read the annotations. The tutorial will grow with input
from experienced analysts and future plans include the addition of
fully interactive `quizzes' (Section~\ref{sec:conclusions}). Despite
the simplicity of the current tutorials, they are already quite
effective, as demonstrated by the crowd-sourcing experiment described
in the next section of this paper.

@@

%\begin{figure}[!ht]
{ \centering \includegraphics[width=15cm]{radialcrowd.pdf}
  \captionof{figure}{a) box plots of 25 crowd-sourced sets of fission
    track counts for first-time student 70 analysts using ROIs defined
    by PV. b) radial plot comparing the paired track counts for the
    first two grains in the dataset (4649 and 4673). The estimate of
    user PV is shown in blue. Analyst AC used a different ROI and is
    not shown.\medskip}
  \label{fig:radialcrowd}
}%\end{figure}

@@

Figure~\ref{fig:radialcrowd} presents the results of a crowd-sourcing
experiment that was carried out at UCL as part of an undergraduate
course in isotope geology. 70~students were asked to create an account
on the geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), the students were asked to
count fission tracks in randomly assigned image stacks of Mount
Dromedary apatite. Each student was required to analyse at least 15
grains, using pre-defined ROIs. In a matter of hours, the students
amassed a dataset of 37,331 fission track counts in 34 grains. This
large dataset was parsed into separate data files --one for each
student-- which they processed during an assessed practical
exercise.\medskip

The raw counts are shown as a series of box plots in
Figure~\ref{fig:radialcrowd}a for the same selection of 25 grains that
were analysed by AC and PV in Section~\ref{sec:private}. The box plots
exhibit significant variability between the grains and between the
students. The inter-grain variability is expected. It reflects the
different sizes of the ROIs, and the variable U-content of the
grains. The inter-student variability is caused by observer bias. This
phenomenon was previously discussed in
Section~\ref{sec:private}. However, the bias is much stronger for the
crowd-sourced dataset than it was for the two `expert' analysts of
Section~\ref{sec:private}. Figure~\ref{fig:radialcrowd}b compares the
42 raw fission track counts of the first two grains on a radial plot,
where
\begin{equation}
  \frac{\mbox{se}(N_1/N_2)}{N_1/N_2}
  \approx
  \sqrt{
    \frac{1}{N_1} + \frac{1}{N_2}
  }
  \label{eq:N12}
\end{equation}

The pairwise ratios of the counts for each analyst are more consistent
than the raw counts of Figure~\ref{fig:crowdsourcing}a.  Students who
over- or undercount the first grain, tend to over- or undercount the
second grain as well. Nevertheless, the results are overdispersed with
respect to the counting uncertainties. The overdispersion is
particularly bad considering that Equation~\ref{eq:N12} ignores the
systematic uncertainty caused by the fact that all analysts counted
exactly the same grains.  The poor consistency of the results reflects
the very limited training received by the students. But despite these
inconsistencies, the opinion of an `expert' (PV, shown in blue) falls
the middle of the estimates obtained by the citizen scientists. This
suggests that the collective wisdom of the entire group of students is
greater than that of any individual analyst.\medskip

>>>>

geochron@home serves four different purposes. It can be used (1) to
count fission tracks in `private mode', i.e. hidden from other users
on the internet; (2) to archive fission track images and counts for
inspection by other users; (3) to create tutorials for new students of
the fission track method; and (4) to generate crowd-sourced fission
track data, by serving randomly selected selections of images to
citizen scientists. We illustrate these four applications with
examples that demonstrate (1) geochron@home's ability to compare and
combine fission track counts for multiple users within a lab group;
(2) the value of the geochron@home archive in the peer review system;
(3) the use of simple tutorials in teaching novice users how to count
fission tracks; and (4) the `wisdom of crowds' in fission track
identification.\medskip

@@

\item Upload the data to the geochron@home platform. geochron@home is
  a Django web-app with a PostgreSQL database.  The database can be
  accessed via a Python API and a (more limited) web-based
  GUI. Accessing the API requires administrator
  privileges. Administrators can create projects, samples and grains;
  download data; and set the access rights of `ordinary' users.
  Projects can be private or public. The source code and installation
  instructions for geochron@home are freely available over GitHub (see
  the Data Availability statement at the end of this paper). This
  allows fission track users to set up their own
  server. Alternatively, fission track labs can upload their data to
  the UCL server by contacting the corresponding author.

\item Analyse the images with a browser-based `virtual fission track
  microscope' powered by the Leaflet library (JavaScript). This
  virtual instrument acts as a front-end to geochron@home. It has a
  simple user interface with controls to zoom, pan and focus in or out
  of the digital image stack. Depending on the permissions granted to
  the user by the administrator, the virtual microsocope offers a
  number of different options.  Entry level `ordinary' users are only
  allowed to count tracks by clicking within a pre-defined `region of
  interest'. In contrast, `superusers' are allowed to define their own
  regions of interest. Once the user is satisfied that they have
  counted all the fission tracks in a particular grain, they can
  submit the results to the server. They are then presented with a new
  set of images until all grains are counted.

@@

The annotations in the current tutorial pages were made by an
experienced fission track analyst (Andrew Carter). This basic tutorial
provides a quick and easy mechanism to train novice users in the art
of fission track analysis. The tutorial pages are in their infancy,
and offer only a limited degree of interactivity. Users can click on
features to read the annotations. The tutorial will grow with input
from experienced analysts and future plans include the addition of
fully interactive `quizzes' (Section~\ref{sec:conclusions}). The
limitations of the current tutorials are apparent in the results of
the crowd-sourcing experiment described in the next section of this
paper.

@@

%\begin{figure}[!ht]
{ \centering \includegraphics[width=\linewidth]{radialcrowd.pdf}
  \captionof{figure}{a) box plots of 25 crowd-sourced sets of fission
    track counts for 69 analysts; b) comparison of grains~23 and 25,
    which were chosen because of their high counts and variability; c)
    radial plot of the pairwise ratios of the fission track counts for
    grains~23 and 25. The results of analyst PV are shown in
    blue.\medskip}
  \label{fig:radialcrowd}
}%\end{figure}

@@

Figure~\ref{fig:radialcrowd} presents the results of a crowd-sourcing
experiment that was carried out at UCL as part of an undergraduate
course in isotope geology. 68~students were asked to create an account
on the geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), the students were asked to
count fission tracks in randomly assigned image stacks of Mount
Dromedary apatite. Each student was required to analyse at least 15
grains, using pre-defined ROIs. In a matter of hours, the students
amassed a dataset of 37,331 fission track counts in 34 grains. This
large dataset was parsed into separate data files --one for each
student-- which they processed during an assessed practical
exercise.\medskip

The raw counts are shown as a series of box plots in
Figure~\ref{fig:radialcrowd}a for the same selection of 25 grains that
were analysed by AC and PV in Section~\ref{sec:private}. The box plots
exhibit significant variability between the grains and between the
students. The inter-grain variability is expected. It reflects the
different sizes of the ROIs, and the variable U-content of the
grains. The inter-student variability is caused by observer bias. This
phenomenon was previously discussed in
Section~\ref{sec:private}. However, the bias is much stronger for the
crowd-sourced dataset than it was for the two `expert' analysts of
Section~\ref{sec:private}. The majority of the student analysts
undercounted the fission tracks by up to a factor of five compared to
the `expert' analyst (PV, shown in blue). The poor consistency of the
results reflects the very limited training received by the
students.\medskip

Grains 23 and 25 have greater than average track densities, and are
particularly dispersed. Plotting the 37 paired counts for each user on
a scatter plot reveals a strong correlation
(Figure~\ref{fig:radialcrowd}b). This means that students who over- or
undercount the first grain, also tend to over- or undercount the
second grain. Even though most students count fewer fission tracks
than PV, the ratio of their counts are similar.
Figure~\ref{fig:radialcrowd}c compares grains 23 and 25 on a radial
plot, where
\begin{equation}
  \frac{\mbox{se}(N_1/N_2)}{N_1/N_2}
  \approx
  \sqrt{
    \frac{1}{N_1} + \frac{1}{N_2}
  }
  \label{eq:N12}
\end{equation}

The tight clustering of the data within the 2-se band confirms that
the pairwise ratios of the counts for each analyst are more consistent
than the raw counts of Figure~\ref{fig:crowdsourcing}a.  Despite the
tremendous scatter of the raw track counts obtained by the citizen
scientists, their pairwise ratio estimates scatter evenly around the
expert opinion. This suggests that the collective wisdom of the entire
group of students is greater than that of any individual
analyst.\medskip
